## 1. Load Required Libraries

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
```

## 2.1 Data Loading

```{r}
# read the csv data
#result = read.csv(here("data", "raw", "results.csv"))
results = read.csv("../data/raw/results.csv")
```

## 2.2 Quick Data Overview

```{r overview}
cat("Dataset shape:", nrow(results), "rows x", ncol(results), "columns\n")
cat("Date range:", min(results$date), "to", max(results$date), "\n")
cat("Unique teams:", length(unique(c(results$home_team, results$away_team))), "\n")
cat("Total matches:", nrow(results), "\n")
```

## 3.1 Filter for Modern Era (1998-2024)

```{r filter-era}
results_clean <- results %>%
  filter(year(date) >= 1998) %>%
  mutate(
    year = year(date),
    month = month(date),
    home_score = as.numeric(home_score),
    away_score = as.numeric(away_score)
  ) %>%
  filter(!is.na(home_score), !is.na(away_score), tournament != "Friendly")%>%
    select(-c(tournament, city, country, neutral, date))
```

## 3.2 filter to only fifa member countries

```{r filter to fifa member countries only}
fifa_members_2024 <- c(
  "Afghanistan", "Albania", "Algeria", "American Samoa", "Andorra", "Angola", 
  "Antigua and Barbuda", "Argentina", "Armenia", "Australia", "Austria", 
  "Azerbaijan", "Bahamas", "Bahrain", "Bangladesh", "Barbados", "Belarus", 
  "Belgium", "Belize", "Benin", "Bermuda", "Bhutan", "Bolivia", 
  "Bosnia and Herzegovina", "Botswana", "Brazil", "British Virgin Islands", 
  "Bulgaria", "Burkina Faso", "Burundi", "Cambodia", "Cameroon", "Canada", 
  "Cape Verde", "Cayman Islands", "Central African Republic", "Chad", "Chile", 
  "China", "Colombia", "Comoros", "Congo", "Congo DR", "Costa Rica", 
  "Croatia", "Cuba", "Curacao", "Cyprus", "Czech Republic", "Denmark", 
  "Djibouti", "Dominica", "Dominican Republic", "Ecuador", "Egypt", 
  "El Salvador", "England", "Equatorial Guinea", "Eritrea", "Estonia", 
  "Eswatini", "Ethiopia", "Faroe Islands", "Fiji", "Finland", "France", 
  "Gabon", "Gambia", "Georgia", "Germany", "Ghana", "Gibraltar", "Greece", 
  "Grenada", "Guatemala", "Guinea", "Guinea-Bissau", "Guyana", "Haiti", 
  "Honduras", "Hong Kong", "Hungary", "Iceland", "India", "Indonesia", 
  "Iran", "Iraq", "Israel", "Italy", "Ivory Coast", "Jamaica", "Japan", 
  "Jordan", "Kazakhstan", "Kenya", "Kosovo", "Kuwait", "Kyrgyzstan", "Laos", 
  "Latvia", "Lebanon", "Lesotho", "Liberia", "Libya", "Liechtenstein", 
  "Lithuania", "Luxembourg", "Macau", "Madagascar", "Malawi", "Malaysia", 
  "Maldives", "Mali", "Malta", "Mauritania", "Mauritius", "Mexico", 
  "Moldova", "Monaco", "Mongolia", "Montenegro", "Morocco", "Mozambique", 
  "Myanmar", "Namibia", "Nepal", "Netherlands", "New Caledonia", "New Zealand", 
  "Nicaragua", "Niger", "Nigeria", "North Korea", "North Macedonia", 
  "Northern Ireland", "Norway", "Oman", "Pakistan", "Palestine", "Panama", 
  "Papua New Guinea", "Paraguay", "Peru", "Philippines", "Poland", "Portugal", 
  "Puerto Rico", "Qatar", "Romania", "Russia", "Rwanda", "Saint Kitts and Nevis", 
  "Saint Lucia", "Saint Vincent and the Grenadines", "Samoa", "San Marino", 
  "Sao Tome and Principe", "Saudi Arabia", "Scotland", "Senegal", "Serbia", 
  "Seychelles", "Sierra Leone", "Singapore", "Slovakia", "Slovenia", 
  "Solomon Islands", "Somalia", "South Africa", "South Korea", "South Sudan", 
  "Spain", "Sri Lanka", "Sudan", "Suriname", "Sweden", "Switzerland", "Syria", 
  "Tahiti", "Tajikistan", "Tanzania", "Thailand", "Timor-Leste", "Togo", 
  "Tonga", "Trinidad and Tobago", "Tunisia", "Turkey", "Turkmenistan", 
  "Turks and Caicos Islands", "Tuvalu", "Uganda", "Ukraine", 
  "United Arab Emirates", "United States", "United States Virgin Islands", 
  "Uruguay", "Uzbekistan", "Vanuatu", "Venezuela", "Vietnam", "Wales", 
  "Yemen", "Zambia", "Zimbabwe"
)

# Filter to FIFA members only
results_fifa_member <- results_clean %>%
  filter(home_team %in% fifa_members_2024, away_team %in% fifa_members_2024)
```

```{r check every column's name}
cat("After filtering, column names are:", for (col in colnames(results_fifa_member)) {
  print(col)
}
)
```

## 3.3 Create Team ID Mapping

```{r team-ids}
all_teams <- sort(unique(c(results_fifa_member$home_team, results_fifa_member$away_team)))
team_mapping <- tibble(
  team = all_teams,
  team_id = seq_along(all_teams)
)

head(team_mapping, 15) %>% kable()
cat("Total teams in modern era:", nrow(team_mapping), "\n")
```

# 3.4 Add Team IDs to Matches

```{r add-ids}
matches <- results_fifa_member %>%
  left_join(team_mapping, by = c("home_team" = "team")) %>%
  rename(home_team_id = team_id) %>%
  left_join(team_mapping, by = c("away_team" = "team")) %>%
  rename(away_team_id = team_id) %>%
  mutate(
    match_id = row_number(),
    total_score = home_score + away_score,
    goal_diff = home_score - away_score,
    home_win = as.integer(home_score > away_score),
    draw = as.integer(home_score == away_score),
    away_win = as.integer(home_score < away_score)
  )

head(matches) %>% kable()
```

## 3.5 Define Eras for Hierarchical Model

```{r define-eras}
matches <- matches %>%
  mutate(
    era = case_when(
      year <= 2002 ~ 1,  # 1998-2002
      year <= 2006 ~ 2,  # 2002-2006
      year <= 2010 ~ 3,  # 2006-2010
      year <= 2014 ~ 4,  # 2010-2014
      year <= 2018 ~ 5,  # 2014-2018
      year <= 2022 ~ 6,  # 2018-2022
      TRUE ~ 7           # 2022-2026 (recent for 2026 prediction)
    ),
    era_label = case_when(
      era == 1 ~ "1998-2002",
      era == 2 ~ "2002-2006",
      era == 3 ~ "2006-2010",
      era == 4 ~ "2010-2014",
      era == 5 ~ "2014-2018",
      era == 6 ~ "2018-2022",
      era == 7 ~ "2022-2026"
    )
  )

matches %>% count(era, era_label) %>% kable()
```

## 3.6 Create Train/Test Split

```{r train-test}
# Use most recent 2024 matches as test set for predictive validation
# Or use random 10% split
test_matches <- matches %>% filter(year > 2022) 
train_matches <- matches %>% anti_join(test_matches, by = "match_id")

cat("Training matches:", nrow(train_matches), "\n")
cat("Test matches:", nrow(test_matches), "\n")
```

# 4. Exploratory Data Analysis

## 4.1 Goal Difference Distribution

```{r}
matches %>%
  mutate(goal_diff = home_score - away_score,
         result = case_when(
           goal_diff > 0 ~ "Home Win",
           goal_diff == 0 ~ "Draw",
           goal_diff < 0 ~ "Away Win"
         )) %>%
  ggplot(aes(x = goal_diff, fill = result)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_fill_manual(values = c("Home Win" = "steelblue",
                               "Draw" = "gray",
                               "Away Win" = "coral")) +
  labs(title = "Goal Difference (Home - Away)",
       x = "Goal Difference",
       y = "Frequency") +
  theme_minimal()
```

## 4.2 Home Advantage

```{r}
home_summary <- matches %>%
  summarise(
    home_win_pct = mean(home_win) * 100,
    draw_pct = mean(draw) * 100,
    away_win_pct = mean(away_win) * 100,
    avg_home_goals = mean(home_score),
    avg_away_goals = mean(away_score)
  )

home_summary %>% kable(digits = 2)

# Visualize
tibble(
  outcome = c("Home Win", "Draw", "Away Win"),
  percentage = c(home_summary$home_win_pct, home_summary$draw_pct, home_summary$away_win_pct)
) %>%
  ggplot(aes(x = outcome, y = percentage, fill = outcome)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), vjust = -0.5) +
  ylim(0, 50) +
  labs(title = "Match Outcome Distribution") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 4.3 Goals Over Time

```{r}
matches %>%
  group_by(year) %>%
  summarise(avg_total_goals = mean(total_score)) %>%
  ggplot(aes(x = year, y = avg_total_goals)) +
  geom_line(color = "steelblue") +
  geom_point() +
  labs(title = "Average Total Goals per Match Over Time", x = "Year", y = "Avg Goals") +
  theme_minimal()
```

## 4.4 Top Teams by Match Count (for reference)

```{r}
team_counts <- matches %>%
  select(team = home_team) %>%
  bind_rows(matches %>% select(team = away_team)) %>%
  count(team, sort = TRUE) %>%
  head(20)

team_counts %>% kable()
```

## 4.5 2026 World Cup Participants

```{r}
# Teams expected in 2026 World Cup (expand as needed)
wc_2026_teams <- c(
  # Auto-qualified hosts
  "United States", "Canada", "Mexico",
  # Major historical teams (add more as qualification progresses)
  "Brazil", "Argentina", "France", "Germany", "Spain", "England",
  "Netherlands", "Italy", "Portugal", "Belgium", "Uruguay",
  "Japan", "South Korea", "Australia"
)

wc_teams_present <- team_mapping %>%
  filter(team %in% wc_2026_teams) %>%
  arrange(team)

wc_teams_present %>% kable()
cat("\nWorld Cup teams with data:", nrow(wc_teams_present), "/", length(wc_2026_teams), "\n")
```

# 5. Save Processed Data

```{r}
# Save team mapping
write_csv(team_mapping, "../data/processed/team_mapping.csv")

# Save training data for Stan
stan_data <- list(
  N = nrow(train_matches),
  T = nrow(team_mapping),
  E = 4,  # number of eras
  home_team = train_matches$home_team_id,
  away_team = train_matches$away_team_id,
  era = train_matches$era,
  home_goals = train_matches$home_goals,
  away_goals = train_matches$away_goals
)

# Save as CSV and RDS for R
write_csv(train_matches, "../data/processed/train_matches.csv")
write_csv(test_matches, "../data/processed/test_matches.csv")
write_rds(stan_data, "../data/processed/stan_data.rds")

message("Data saved to data/processed/")
```

# 6. Summary for Report

```{r}
summary_stats <- list(
  total_matches = nrow(matches),
  train_matches = nrow(train_matches),
  test_matches = nrow(test_matches),
  unique_teams = nrow(team_mapping),
  date_range = paste(min(matches$date), "to", max(matches$date)),
  avg_home_goals = round(mean(matches$home_goals), 3),
  avg_away_goals = round(mean(matches$away_goals), 3),
  home_win_rate = round(mean(matches$home_win), 3),
  matches_by_era = matches %>% count(era, era_label) %>% select(-era) %>% deframe() %>% as.list()
)

print(summary_stats)
```
